---
title: "Using Legacy Newhall v1.6.1 Routines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Legacy Newhall v1.6.1 Input/Output Routines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates setting a custom `jNSMR.JAR_SUFFIX` to force use of the legacy Java Newhall Model version 1.6.1 (released 2016/02/10). See official download here: https://www.nrcs.usda.gov/wps/portal/nrcs/detail/?cid=nrcs142p2_053559.

## Copyright Information

 > Newhall 1.6.1, Copyright (C) 2010-2011 United States Department of Agriculture - Natural Resources Conservation Service, Penn State University Center for Environmental Informatics All rights reserved.

```{r setup}
options(jNSMR.JAR_SUFFIX = "-1.6.1")
library(jNSMR)
```

### Run the BASIC simulation model

Create a _NewhallDataset_ with `xml_NewhallDataset()` or `NewhallDataset()` and associated _NewhallDatasetMetadata_ with `NewhallDatasetMetadata()`.

```{r}
library(jNSMR)

# read single-station XML file (NewhallDataset + NewhallDatasetMetadata)
input_xml_file <- system.file("extdata/WILLIAMSPORT_1930_1930_input.xml", package = "jNSMR")[1]
input_xml <- xml_NewhallDataset(input_xml_file)

# or specify inputs directly to the constructor
input_direct <- NewhallDataset(stationName = "WILLIAMSPORT",
                               country = "US",
                               lat = 41.24,
                               lon = -76.92,
                               elev = 158.0,
                               allPrecipsDbl = c(44.2,40.39,113.54,96.77,95.0,98.55,
                                                 66.04,13.46,54.86,6.35,17.53,56.39),
                               allAirTempsDbl = c(-2.17,0.89,3.72,9.11,16.28,21.11,
                                                  22.83,21.94,19.78,10.5,5.33,-1.06),
                               pdbegin = 1930,
                               pdend = 1930,
                               smcsawc = 200.0)

# set minimal _NewhallDatasetMetadata_ for our constructed _NewhallDataset_
input_direct$setMetadata(NewhallDatasetMetadata(stationName = "WILLIAMSPORT"))
```

### Run single model from file or R object

```{r}
# run single model (from XML file--only with v1.6.1 JAR)
output_xml <- newhall_simulation(input_xml)

# run single model (from direct input)
output <- newhall_simulation(input_direct)
```

### Access references to Java objects containing results

Here we demonstrate how to use the `NewhallResults` object from `newhall_simulation()` to get various formatted statistics, calendars and other output from a single run of the model.

#### Moisture and temperature regime statistics

```{r}
cat(output$getFormattedStatistics())
```

#### Moisture calendar

```{r}
cat(output$getFormattedMoistureCalendar())
```

#### Temperature calendar

```{r}
cat(output$getFormattedTemperatureCalendar())
```

### Support for Newhall XML format for input and output

`newhall_XMLStringResultsExporter()` returns a character vector with XML based on an input and an output object.

```{r}
# inspect XML result format
cat(newhall_XMLStringResultsExporter(input_xml, output))
```

### XML File Output

This writes the "direct" input _NewhallDataset_ (`input_direct`) and result _NewhallResult_ (`output`) to an XML file with format as above. 

```{r}
# write XML results to file
newhall_XMLResultsExporter(dataset = input_direct,
                           result = output,
                           pathname = "misc/WILLIAMSPORT_1930_1930_export.xml")
```

### Graphical User Interface for individual runs via _newhall-1.6.1.jar_

`newhall_GUI()` is convenience method to open the Newhall GUI (single runs in v1.6.1 only) with the R package.

```{r, eval=FALSE}
newhall_GUI()
```
