---
title: "Applying Newhall to PRISM (4km) 30-year Normals for Continental United States"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applying Newhall to PRISM (4km) 30-year Normals for Continental United States}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

```{r setup}
library(jNSMR)
library(prism)
library(terra)

# run jNSMR on prism monthly 30yr normals

# path to PRISM data
PRISM_PATH <- tempdir() #"~/Geodata/PRISM"
prism::prism_set_dl_dir(PRISM_PATH)

# PRISM resolution; either 4km or 800m
RESOLUTION <- "4km"
```

Download PRISM data (if needed)

```{r}
# download 4km or 800m data
.download_month <- function(i, resolution = c("4km","800m"), annual = FALSE, keepZip = TRUE) {
  resolution <- match.arg(resolution, c("4km","800m"))
  if(i == 1)
    annual = TRUE
  prism::get_prism_normals(mon = i, type = "ppt",
                    resolution = resolution,
                    annual = annual, keepZip = keepZip)
  prism::get_prism_normals(mon = i, type = "tmean",
                    resolution = resolution,
                    annual = annual, keepZip = keepZip)
  TRUE
}

bilfile <- list.files(PRISM_PATH, "\\.bil$", recursive = TRUE)

# only download the PRISM data if it can't be found in PRISM_PATH
if (length(bilfile) == 0) {
  
  # download by month
  res <- lapply(1:12, .download_month, resolution=RESOLUTION)
  
  # get the monthly files only (hardcoded for 30yr normals for now)
  bilfile <- list.files(PRISM_PATH, "\\.bil$", recursive = TRUE)
}

# capture file name components
monthly.pattern <- ".*PRISM_([a-z]+)_(30yr_normal)_(\\d+[a-z]+)M[23]_(\\d{2})_bil\\.bil"
bilfile_sub <- bilfile[grep(monthly.pattern, bilfile)]
bilmonth <- strsplit(gsub(monthly.pattern,  "\\1;\\2;\\3;\\4", bilfile_sub), ";")
bil <- data.frame(bilfile_sub, do.call('rbind', bilmonth))
colnames(bil) <- c("bilfile","variable","product","resolution","month")
bilsub <- subset(bil, resolution == RESOLUTION)
```

Create PRISM raster stack and extract POINT (longlat) locations of cells

```{r}
# stack rasters, create matrix of values and XY (longlat) locations as data.frame
prism_stack <- terra::rast(file.path(PRISM_PATH, bilsub$bilfile))
names(prism_stack) <- paste0(bilsub$variable, bilsub$month)
cx <- terra::xyFromCell(prism_stack, 1:ncell(prism_stack))
prism_stack$lonDD <- cx[,1]
prism_stack$latDD <- cx[,2]
prism_frame <- as.data.frame(terra::as.points(prism_stack))
```

Prepare jNSM batch input (CSV) file

```{r}
# fix column names for jNSM style batch input format
monthnames <- month.abb[sapply(names(prism_frame), function(x) as.integer(gsub("([a-z])", "", x)))]
newnames <- gsub("tmean", "t", names(prism_frame))
newnames <- gsub("ppt", "p", newnames)
newnames <- gsub("\\d", "", newnames)
newnames <- paste0(newnames, ifelse(is.na(monthnames), "", monthnames))
names(prism_frame) <- newnames

# PRISM info
prism_frame$stationName <- 1:nrow(prism_frame)
prism_frame$awc <- 200
prism_frame$maatmast <- 1.2
prism_frame$pdType <- "Normal"
prism_frame$pdStartYr <- 1981
prism_frame$pdEndYr <- 2010
prism_frame$cntryCode <- "US"

# boilerplate minimum metadata -- these are not used by the algorithm
prism_frame$netType <- ""
prism_frame$elev <- -9999
prism_frame$stProvCode <- ""
prism_frame$notes <- ""
prism_frame$stationID <- 1:nrow(prism_frame)

# make test batch of prism point data values
test_set <- prism_frame[round(runif(20000, 1, nrow(prism_frame))),]
```

## Run simulations

```{r}
# run simulations
res <- jNSMR::newhall_batch(test_set) 
```

## Results

Create terra SpatVector containing points and results

```{r}
newsp <- terra::vect(cbind(test_set[,c('lonDD','latDD')], res), geom = c("lonDD","latDD"), crs="EPSG:4326")
```

Temperature Regimes

```{r}
plot(newsp, col=factor(newsp$temperatureRegime))
```

Moisture Regimes

```{r}
plot(newsp, col=factor(newsp$moistureRegime))
```

Moisture Regime Subdivision 1

```{r}
plot(newsp, col=factor(newsp$regimeSubdivision1))
```

Moisture Regime Subdivision 2

```{r}
plot(newsp, col=factor(newsp$regimeSubdivision2))
```

Moisture regimes (Xeric, Aridic, Ustic, Udic)

```{r}
newspsub <- subset(newsp, newsp$moistureRegime %in% c("Aridic","Udic","Ustic","Xeric"))
legnames <- factor(paste(newspsub$moistureRegime))
nlegnames <- unique(levels(legnames))
nlegcolors <- c("RED","BLUE","ORANGE","YELLOW")
plot(newspsub, col = nlegcolors[match(legnames, nlegnames)], pch=".", cex=5)
legend("topright", legend = nlegnames, pch=".", fill = nlegcolors)
```

Moisture regime subdivisions within Aridic and Xeric

```{r}
table(newspsub$regimeSubdivision1)
newspsub2 <- subset(newsp, newsp$moistureRegime %in% c("Aridic","Xeric"))
legnames <- factor(paste(newspsub2$regimeSubdivision1, newspsub2$moistureRegime))
nlegnames <- unique(levels(legnames))
nlegcolors <- c("ORANGE","RED","TAN","YELLOW","SALMON")
plot(newspsub2, col = nlegcolors[match(legnames, nlegnames)], pch=".", cex=5)
legend("bottomleft", legend = nlegnames, pch=".", fill = nlegcolors)
```

```{r, echo=FALSE}
unlink(PRISM_PATH, recursive = TRUE)
```
